<Concise statement of the problem you intended to solve and main achievements (no more than one A4 page)>

\vspace{10pt}

As the commercial use of Unmanned Aerial Systems (UASs) for tasks such as surveillance, infrastructure maintenance, and goods delivery increases, the density of urban air traffic will exceed the capacity of traditional human control centers. To ensure safety and efficiency at scale, an automated unmanned approach to UAS Traffic Management (UTM) is required, in which Airspace Service Providers (ASPs) nogotiate flight plans with operators without continous human oversight.\par

\vspace{10pt}

This project addresses that challenge by Treating urban airspace managment as a Multi-Agent Path Finding (MAPF) problem and evaluating a range of management polices through simulation.A custom discrete event simulaiton framework was developed in Python using SimPy, capable of modelling one or more ASPs serving dynamic flight-plan requests accross a configurable urban grid.  \par

Eight policies were implemented and compared, ranging from uninformed search baselines (BFS, DFS) and a naive direct-path policy that ignores obstacles and other agents to provide a lower-bound reference on achievable flight time, through single-agent informed search (A*) and a variant aware of dynamic occupancy (Occupancy A*), to prioritised and cooperative coordination strategies (Prioritised A*, Cooperative A*, and Windowed Hierarchical Cooperative A*). Each policy was evaluated under multiple demand levels and reservation configurations, with 5 independent random seeds per condition to support reproducible statistical comparison. \par

\vspace{10pt}

The evaluation produced three principal findings. First, the segment-based reservation mechanism eliminates UAV collisions across the policies tested. However, it comes at a measurable cost in completion rate and average delivery time. Second, prioritised and cooperative methods outperform uninformed search baselines on safety and efficiency, but the dominant comparison between them depends on the demand level and reservation horizon rather than on the choice of search algorithm in isolation. Third, the completability of prioritised time-expanded search is bounded by the reservation horizon rather than by available computation. \par

\vspace{10pt}

The results indicate that the principal driver of safe and efficient ASP-mediated UAS coordination is the presence of an explicit time-aware coordination mechanism rather than the sophistication of the underlying single-agent planner. Reservation-based coordination is shown to be a practical and tractable alternative to full space-time MAPF for ASP-mediated UTM. Directions for future work include the integration of  coupledsearch methods such as Conflict Based Search, the evaluation of tactical reactive avoidance methods, and the extension of the framework to three-dimensional airspace.

