\chapter{Methodology}
\label{chapter2}

% <Everything that comes under the `Methodology' criterion in the mark scheme should be described in one, or possibly more than one, chapter(s).>

This chapter describes the design and methodology of the simulation framework developed in this project. Section 2.1 outlines the requirements and design goals that motivated the frameowrk. Section 2.2 presents the archiecture. Section 2.3 describes the model with focus on the descrete event approach. Section 2.4 describes the management polies and strategy pattern through which they are composed. Section 2.5 describes the segment-based reservation mechanism that forms the central experimental variable in the evaluation. Finally, section 2.6 documents the project managment approch adopted during development. \par

\section{Requirements and design goals}

The literature reviewed in chapter 1 identifies a gap in the comparative empirical evaluation of management polices for ASP-mediated UAS Traffic management (UTM), closing that gap requires a simualtion framework that is sufficialtly faithful to the operational context of UTM to make its results meaningful, and sufficiently flexible to support controlled comparison across a range of policies and operating conditons. This section sets out the requirements created with these considerations in mind. Which together infrom the design choices documented in the remainder of the chapter. 

The functional requirements are as follows. The framework must model an urban airspace as a network of independent Airspace Service providers (ASPs), each controlling thier defined area and handling the handover of traffic to its neighbouring controllers. It must process flight plan requests dynamically during execution rather then as a fixed batch known in advance (to reflect the real world) it must support multiple management from MAPF reviewed in the chapter above. it must permit policies to be tested in the same envrionments and configurations. it must record sufficent data on each flight and simulation to support the saftey, efficency and coordination metrics described. 

The non functional requirements are equally importatnt. The framework must be deteminsitic given a fixed random seed, so that experimental results are reporducible. It must be modular enough to permit new polices to be added without modifying the siulation core, in support of both the comparitive evaluations undertaken in this work and any future extensions. It must provide configeration through external files rather then paremeters so experimental conditions can be varied without recompliation. It must complete experiments within a resanable time on comodity hardware as it will be ran many time in many configeration.

From these requirements they derive several other design goals that reoccur throughout the rest of this chapter. The simulation must be discrete event driven rather than time stepped. Many considerations motivated this choice. First, performance: a time-stepped simulation processes every agent (UAV) at every clock tick regardless of whether there state has changed, while a discrete-event simulation processes a UAV only when an event affects it, which scales considerably better as the number of UAVs and the length of the simulated horizon grow. Secondly, precision in collision detection: continuous time geometric collision detection over analytic trajectory segments yields the exact moment and distance of closest approach, while a fixed-tick simulation can miss close-approach events that occur entirely within a single time step unless the step is made impractically small. Third, reproducibility: a discrete-event simulation is fully determined by its random seed and configuration, whereas a time-stepped simulation introduces the time step itself as a hidden parameter whose value can affect experimental results. SimPy \cite{simpy2024} is the natural choice for this paradigm in Python.

These requirements and design goals together motivate the architecture presented in Section 2.2, the simulation model described in Section 2.3, the policy abstraction documented in Section 2.4, the reservation mechanism developed in Section 2.5, and the project management approach summarised in Section 2.6.

\section{requirements and design goals}

The simulation frameowkr is organsed as a layered modular architecture in which each layer has a clearly defined responsablity, relying on the layer below. the arcihtecture illisturated in <architectrue figure>, comprises of these key layers: configeration, world structure, ASP, agents, planning, collision avoidance and finaly metric collection.

<figure>

At the top of the archiecture is a single simulation object. It serves as the orchestrator and it is constructed from the configuation file. It is responsable for instatiating the SimPu envrionment, the metrics collector, creating the plannng polcies and the world manager. Once instantiated it deligaes teh running of the simulation event loop to a SimPy event manager after this its only responsable for recording the final metrics to a file on completion.

The world manger owns the world strucuter and manger the inter-airspace coordinatino. It loads a world manifest that specifies the layout of airspaces, the position and orientation of each, and the connectivity betwee them. From the given manifest it also instatiates an Airspace object per sector. Once the airspaces have been created they are represented as nodes in a graph and the gates are added as edges. This is what is used when a UAV requres a route that crosses airspace boundires and it returns a sequesnce of airspaces and gates that the local planner within each airspace uses to get the UAV from its current position to a gate. 

Each Airspace repersents one sector of the simulated airspace and it owns these static elements within; the grid mpa that represnts teh sectors geomatry, traesablity, depots and pbstalces; 